---
title: "Mutational burden in normal human small intestine"
author: "Yichen Wang"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=70), tidy=TRUE,warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = '/Users/yw2/OneDrive\ -\ University\ of\ Cambridge/PhD/code/small_bowel/')
```

This analysis aims to assess 1) the rate of SBS and ID mutations, 2) the presence and the rate of mutaional burden of mutational signatures in normal small intestine.

We build linear mixed-effect models to estimate mutation rates. Model parameters encoded are: age, biopsy region (duodenum, jejunum, ileum) , celieac condition (having a coeliac history or not) and  patient ID. We have corrected for coverage by normalising ID and SBS mutational burden according to the sensitivity of detection in each samples before modelling, therfore we do not include it as a parameter for modelling.

## 1. Overview of the dataset


```{r loading, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(nlme)
library(tidyverse)

# Load data
data=read.csv('./data/Supplementary_Table4_crypt_summary.csv',header=T,stringsAsFactors = F)
exposure_matrix_crypts = read.table("./data/signatures/exposure_matrix_crypts.txt",header=T,stringsAsFactors = F)
  

exposure_matrix_crypts = exposure_matrix_crypts[data$sample,]
data$SBS1 <- data$sbs_count_adj*exposure_matrix_crypts$SBS1
data$SBS5 <- data$sbs_count_adj*exposure_matrix_crypts$SBS5
data$SBS2 <- data$sbs_count_adj*exposure_matrix_crypts$SBS2
data$SBS13 <- data$sbs_count_adj*exposure_matrix_crypts$SBS13
data$SBS18 <- data$sbs_count_adj*exposure_matrix_crypts$SBS18
data$SBS88 <- data$sbs_count_adj*exposure_matrix_crypts$SBS88
data$SBS35 <- data$sbs_count_adj*exposure_matrix_crypts$SBS35
data$SBS17b <- data$sbs_count_adj*exposure_matrix_crypts$SBS17b
data$SBS40 <- data$sbs_count_adj*exposure_matrix_crypts$SBS40
data$SBS41 <- data$sbs_count_adj*exposure_matrix_crypts$SBS41

# Exclude Brunner's glands
data<-data[which(data$ref=='Crypt'),]
table(data$patient)
dim(data)


df_regression = data
# Only includes samples with >15 coverage
df_regression<-df_regression[which(df_regression$coverage>15),]
# Exclude the individual with substantial mutations from chemotherapy 
df_regression<-df_regression[which(df_regression$patient != 'PD43853'),]

# Exclude the two unusual cases
df_regression<-df_regression[! df_regression$sample %in% c('PD46565c_lo0009','PD43851j_P52_DDM_E2'),]

df_regression$condition <- as.factor(df_regression$condition)
df_regression$condition <- relevel(df_regression$condition, ref = "Normal")
rownames(df_regression)=df_regression$sample
dim(df_regression)

```


```{r}
# Mean coverage
mean(data$coverage)

# Load SBS matrix
sbs<-read.table('./data/mutation_matrices/sbs_mapped_to_branches.txt',check.names = F)
# Exclude Brunner's glands
sbs=sbs[!(rownames(sbs) %in% c('PD43851_1','PD43851_4','PD43851_14')),]
# Total SBS mutations
sum(sbs)


# Load Indel matrix
indel<-read.table('./data/mutation_matrices/indel_mapped_to_branches.txt',check.names = F)
# Exclude Brunner's glands
indel=indel[!(rownames(indel) %in% c('PD43851_1','PD43851_4','PD43851_14')),]
# Total ID mutations
sum(indel)
```

## 2. Hypothesis testing of linear mixed effects models
### Substitutions

First, we set up a linear mixed model with age as a fixed effect and patient as random effect.
```{r, warning=FALSE}
lmm.sbs.age <- lme(sbs_count_adj ~ age , random=list(patient = pdDiag(form = ~ age -1)), data = df_regression, method = 'ML') 
```

Adding coeliac history as a fixed effect improves the fitness of the model.
```{r, warning=FALSE}
# Include coeliac history as a fixed effect does not improve the fitness of model.
lmm.sbs.coeliac.burden <- lme( sbs_count_adj  ~ age + condition, 
                               random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.burden,lmm.sbs.age)
```

Coeliac history does not affect between-patient heterogeneity of single-base substitution burdens.
```{r, warning=FALSE}
# Whether coeliac history affect inter-patient variation of single base substituions
lmm.sbs.coeliac.var.inter <- lme( sbs_count_adj  ~ age + condition, random = list(patient = pdDiag(form = ~ age+condition-1)),  data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.var.inter, lmm.sbs.coeliac.burden)

# Whether coeliac history affect within-patient variation of single base substituions
lmm.sbs.coeliac.var.intra <- lme( sbs_count_adj  ~ age + condition, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.var.intra, lmm.sbs.coeliac.burden)
```

Although the p-value from ANNOVA test for heteroscedasticity is significant, after careful investigation into the raw data, this variation was largely introduced by one single patient, PD46565. After removing this patient, distinguishing the two groups does not provide a better fit. Therefore we don't think coeliac history will affect within-patient variation of single-base substitution burdens.

```{r, warning=FALSE}

lmm.sbs.coeliac.burden <- lme(sbs_count_adj ~ age + condition, random=list(patient = pdDiag(form = ~ age -1)), data = df_regression[which(df_regression$patient != 'PD46565'),], method = 'ML') 
lmm.sbs.coeliac.var.intra <- lme(sbs_count_adj  ~ age + condition, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.sbs.coeliac.var.intra, lmm.sbs.coeliac.burden)

```

Biopsy region does not affect between-patient heterogeneity and heteroscedasticity. 
```{r warning=FALSE}
lmm.sbs.coeliac.burden <- lme(sbs_count_adj ~ age + condition, random=list(patient = pdDiag(form = ~ age -1)), data = df_regression, method = 'ML') 
# Whether biopsy region affect inter-patient variation of single base substituions
lmm.sbs.region.var.inter<- lme( sbs_count_adj  ~ age + condition, 
                                random = list(patient = pdDiag(form = ~ age+region -1)), data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.burden,lmm.sbs.region.var.inter)

# Whether biopsy region affect within-patient variation of single base substituions
lmm.sbs.region.var.intra<- lme( sbs_count_adj  ~ age + condition, 
                                random = list(patient = pdDiag(form = ~ age -1)), weights = varIdent(form = ~1 | region), data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.burden,lmm.sbs.region.var.intra)
```


Allowing different mutation rate for duodenum/jejunum/ileum improves the fitness of the model.
```{r warning=FALSE}
lmm.sbs.region.condition.burden<- lme( sbs_count_adj  ~ age:region + condition, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")
lmm.sbs.condition.burden <- lme(sbs_count_adj ~ age + condition, 
                   random=list(patient = pdDiag(form = ~ age -1)), data = df_regression, method = 'ML') 
anova(lmm.sbs.region.condition.burden,lmm.sbs.condition.burden)
summary(lmm.sbs.region.condition.burden)
# remove coeliac condition from the model as the coefficient is now insignificinat
lmm.sbs.region.burden<- lme( sbs_count_adj  ~ age:region , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

```


Therefore, our final model for SBS burden includes both age and biopsy region as fixed effects, and patient as a random effect.
```{r SBS_plot, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lmm.sbs.region.burden
summary(lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")


tmp=data.frame(condition=c(16,3))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = sbs_count_adj))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age:regionDuodenum',],colour='#F8766D')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'], ymax = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','upper']),fill='#F8766D', alpha = 0.1)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age:regionIleum',],colour='#00B81F')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionIleum',]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionIleum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','upper']), fill='#00B81F',alpha = 0.1)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age:regionJejunum',],colour='#00A5FF')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','upper']), fill='#00A5FF',alpha = 0.1)+
  labs(y='Substitutions / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

### Indels
As above, we set up a linear mixed model for Indels with age as a fixed effect and patient as random effect.
```{r warning=FALSE}
lmm.id.age <- lme(indel_count_adj ~ age, random=list(patient = pdDiag(form = ~ age-1)), data = df_regression, method = 'ML') 
```

Adding seperate mutaion rate for duodenum/jejunum/ileum significantly improves the fitness of the model.
```{r region indel, warning=FALSE}
# Whether biopsy regions affect Indel burden via burden
lmm.id.region.burden <- lme(indel_count_adj  ~ age:region, 
                           random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression, method = "ML")
anova(lmm.id.region.burden,lmm.id.age)
```


No significant inter-patient variation and heteroscedasticity is caused by different biopsy sections.
```{r}
# Biopsy region does not affect between-patient variation of Indels
lmm.id.region.burden.var.inter <- lme( indel_count_adj  ~ age:region , 
                                      random = list(patient = pdDiag(form = ~ age+region-1)),  data = df_regression, method = "ML")
anova(lmm.id.region.burden.var.inter, lmm.id.region.burden)

# Biopsy region does not affect within-patient variation of Indels.The ANNOVA test result is not significant after removing possible outliers.
lmm.id.region.burden.var.intra <- lme( indel_count_adj  ~ age:region , 
                                      random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | region ), data = df_regression, method = "ML")
anova(lmm.id.region.burden.var.intra, lmm.id.region.burden)


lmm.id.region.burden <- lme(indel_count_adj  ~ age:region, 
                           random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
lmm.id.region.burden.var.intra <- lme( indel_count_adj  ~ age:region , 
                                      random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | region ), data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.id.region.burden.var.intra, lmm.id.region.burden)

```


Coeliac history affects Indel burdens and intra-patient variation.
```{r coeliac indel, warning=FALSE}
# Coeliac history affects Indel burdens
lmm.id.region.burden <- lme(indel_count_adj  ~ age:region , 
                           random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression, method = "ML")
lmm.id.region.burden.coeliac.burden <- lme(indel_count_adj  ~ age:region +condition, 
                                           random = list(patient = pdDiag(form = ~ age-1)) ,data = df_regression, method = "ML")
anova(lmm.id.region.burden.coeliac.burden, lmm.id.region.burden)

# Coeliac history does not affect inter-patient variation of Indels (the effect no longer exist after removing one outlier.)
lmm.id.region.burden.coeliac.var.inter <- lme( indel_count_adj  ~ age:region +condition, 
                                              random = list(patient = pdDiag(form = ~ age+condition-1)),  weights = varIdent(form = ~1 | region ), data = df_regression, method = "ML")
lmm.id.region.burden.coeliac.burden <- lme(indel_count_adj  ~ age:region +condition, 
                                           random = list(patient = pdDiag(form = ~ age-1)) ,data = df_regression, method = "ML")
anova(lmm.id.region.burden.coeliac.var.inter, lmm.id.region.burden.coeliac.burden)

lmm.id.region.burden.coeliac.var.inter <- lme( indel_count_adj  ~ age:region +condition, 
                                              random = list(patient = pdDiag(form = ~ age+condition-1)),  weights = varIdent(form = ~1 | region ), data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
lmm.id.region.burden.coeliac.burden <- lme(indel_count_adj  ~ age:region +condition, 
                                           random = list(patient = pdDiag(form = ~ age-1)) ,data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.id.region.burden.coeliac.var.inter, lmm.id.region.burden.coeliac.burden)


# Coeliac history affects intra-patient variation of Indels. The effect exist after removing possible outliers.
lmm.id.region.burden.coeliac.burden.var.intra <- lme( indel_count_adj  ~ age:region +condition, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression, method = "ML")
lmm.id.region.burden.coeliac.burden <- lme(indel_count_adj  ~ age:region +condition, 
                                           random = list(patient = pdDiag(form = ~ age-1)) ,data = df_regression, method = "ML")
anova(lmm.id.region.burden.coeliac.burden.var.intra, lmm.id.region.burden.coeliac.burden)

lmm.id.region.burden.coeliac.burden.var.intra <- lme( indel_count_adj  ~ age:region +condition, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
lmm.id.region.burden.coeliac.burden <- lme(indel_count_adj  ~ age:region +condition, 
                                           random = list(patient = pdDiag(form = ~ age-1)) ,data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.id.region.burden.coeliac.burden.var.intra, lmm.id.region.burden.coeliac.burden)
```


Therefore, our final model for ID burden includes age, biopsy region and coeliac condition as fixed effects, patient as a random effect, and use different covariance matrix for individual with/without a coeliac history.

```{r Indel_plot, fig.width = 6, fig.height=3, warning=FALSE}
lmm.id.region.burden.coeliac.burden.var.intra <- lme( indel_count_adj  ~ age:region +condition, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression, method = "ML")

lmm <-lmm.id.region.burden.coeliac.burden.var.intra
summary(lmm)
fixed.m1 <- data.frame(fixef(lmm))


tmp=data.frame(condition=c(16,3))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = indel_count_adj))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age:regionDuodenum',],colour='#F8766D')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'], ymax = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','upper']),fill='#F8766D', alpha = 0.1)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age:regionIleum',],colour='#00B81F')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','lower'], ymax = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','upper']), fill='#00B81F',alpha = 0.1)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age:regionJejunum',],colour='#00A5FF')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','lower'], ymax = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','upper']), fill='#00A5FF',alpha = 0.1)+
  labs(y='Indels / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```


## 3. Mosaic plot for all signatures
```{r Mosaic_plot,fig.width = 10, fig.height=4}
df=exposure_matrix_crypts[rownames(exposure_matrix_crypts) %in% data$sample,]
df$patient = rownames(df)
df$age =data$age
df$patient = fct_reorder(df$patient, df$age, min)

test <- gather(df, E1, E2, -patient, -age)
test$E1<-factor(test$E1,levels = c('SBS1','SBS5','SBS18','SBS2','SBS13','SBS88','SBS35','SBS40','SBS41','SBS17b'))
test$group = 'Normal'
test$group[test$patient %in% data$sample[data$condition == 'Coeliac']] = 'Coeliac'

sig_order=1:10
names(sig_order)=c("SBS1","SBS5","SBS18","SBS2","SBS13","SBS88","SBS35","SBS40",'SBS41',"SBS17b")
final_sigs=t(read.table("./data/signatures/final_sigs.txt",check.names=FALSE))

getPalette = colorRampPalette(brewer.pal(8, "Set3"))
all_cols=getPalette(8)
all_cols=c(all_cols,"firebrick","magenta")
final_sigs = final_sigs[names(sort(sig_order[rownames(final_sigs)])),]
names(all_cols)=rownames(final_sigs)

ggplot(test, aes(x = patient, y = E2, fill = E1)) + geom_bar(stat = "identity") +scale_fill_manual(values=all_cols)+
  theme_bw()+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(), panel.grid=element_blank(),panel.border=element_blank())+
  labs(x = "Samples", y = "Proportion", fill = "Signatures")+facet_grid(cols = vars(group),scales = "free_x", space = "free")

```

## 4. The presence of APOBEC signatures SBS2/13 in the small intestine
```{r small_bowel_vs_colon, fig.width = 2.5, fig.height = 3}
exposure_matrix_all = read.table("../../small_bowel/data/sigs/exposure_matrix_with_colon.txt",header=T,stringsAsFactors = F)
input_for_hdp<-read.table('../../small_bowel/data/input_for_hdp.txt',check.names = F)
input_for_hdp=input_for_hdp[apply(input_for_hdp,1,sum)>50,]
input_for_hdp=input_for_hdp[!(rownames(input_for_hdp) %in% c('PD43851_1','PD43851_4','PD43851_14')),]

apobec_table = data.frame(matrix(ncol = 2, nrow = length(unique(data$patient))))
rownames(apobec_table)=unique(data$patient)
colnames(apobec_table)=c('Positive','Negative')
for (patient in unique(data$patient)){
  apobec_table[patient,'Positive']=sum(sum(data[data$patient==patient,'SBS2']+data[data$patient==patient,'SBS13']>0.05*data[data$patient==patient,'sbs_count_adj']))
  apobec_table[patient,'Negative']=sum(data$patient==patient)-apobec_table[patient,'Positive']
}

APOBEC_matrix_all=exposure_matrix_all[,c('SBS2','SBS13')]
APOBEC_matrix_all$sum = APOBEC_matrix_all$SBS2 + APOBEC_matrix_all$SBS13
APOBEC_matrix_all$sum[APOBEC_matrix_all$sum<0.05] = 0
APOBEC_matrix_all$group = 'Colon'
APOBEC_matrix_all$group[rownames(exposure_matrix_all) %in% rownames(input_for_hdp)] = 'Small intestine'
APOBEC_matrix_all$branch = rownames(APOBEC_matrix_all)

df=data.frame(matrix(ncol = 3, nrow = 2))
colnames(df) <- c('region', 'pos', 'neg')
df$region = c('Small intestine','Colon')
df$pos = c(sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Small intestine'] > 0),sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Colon'] > 0))
df$neg = c(sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Small intestine'] == 0),sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Colon'] == 0))


df[1,2:3] = df[1,2:3]/sum(df[1,2:3])
df[2,2:3] = df[2,2:3]/sum(df[2,2:3])

test <- gather(df, E1, E2, -region) 
test$E1 <- factor(test$E1,levels = c('pos','neg'))


ggplot(test[test$E1=='pos',]) +
  geom_bar(aes(x = region, y = E2, fill = E1),
           stat = "identity") +scale_fill_brewer(palette="Paired",labels=c('APOBEC positive','APOBEC negative'))+theme_bw()+theme(panel.grid=element_blank(),legend.position="none",legend.title=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  labs(x = "Location", y = "APOBEC frequency")

# aa=chisq.test(rbind(c(68,417-68),c(6,1075-6)))
# aa=chisq.test(rbind(c(37,417-37),c(3,1066-3)))
```

## 5. Mutational burden for common mutational signatures in the normal small intestine
We looked into the accumulation of mutational buren across the corhort for SBS1, SBS5, SBS18 and SBS2/13. Other sporadic signatures are too rare for conducting meaningful statistical analysis.

### SBS1

The accumulation of SBS1 burden is asscoicated with age and differs in diifferent biopsy regions.
```{r SBS1, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS1  ~ age -1, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Biopsy region affects SBS1 burden
lmm.region.burden <- lme(SBS1 ~ age:region -1, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
summary(lmm.region.burden)
# Coeliac history affects SBS1 burden, therefore we need to estimate SBS mutational rate seperately
lmm.coeliac.burden <- lme(SBS1 ~ age:region + condition -1 , random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm.region.burden)


lmm.region.burden <- lme(SBS1 ~ age:region -1, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression[which(df_regression$condition == 'Normal'),], method = "ML")

lmm <- lmm.region.burden
fixed.m1 <- data.frame(fixef(lmm))
summary(lmm)
cor(df_regression$SBS1[which(df_regression$condition == 'Normal')],df_regression$age[which(df_regression$condition == 'Normal')], method="pearson")

tmp=data.frame(condition=c(16,3))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = SBS1))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = 0, slope =  fixed.m1['age:regionDuodenum',],colour='#F8766D')+
  geom_ribbon(aes(ymin = age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'], ymax = age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','upper']),fill='#F8766D', alpha = 0.1)+
  geom_abline(intercept = 0, slope =  fixed.m1['age:regionIleum',],colour='#00B81F')+
  geom_ribbon(aes(ymin = age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','lower'], ymax = age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','upper']), fill='#00B81F',alpha = 0.1)+
  geom_abline(intercept = 0, slope =  fixed.m1['age:regionJejunum',],colour='#00A5FF')+
  geom_ribbon(aes(ymin = age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','lower'], ymax = age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','upper']), fill='#00A5FF',alpha = 0.1)+
  labs(y='SBS1 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```


### SBS5

The accumulation of SBS5 burden is asscoicated with age, but shows no difference across the three sections.
```{r SBS5, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS5  ~ age , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Coeliac history does not affect SBS5 burden
lmm.coeliac.burden <- lme(SBS5 ~ age + condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS5 burden
lmm.region.burden <- lme(SBS5 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS5,df_regression$age, method="pearson")

tmp=data.frame(condition=c(16,3))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = SBS5))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age',])+
  geom_ribbon(aes(ymin = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','lower'], ymax = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','upper']), alpha = 0.1)+
  labs(y='SBS5 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))

```

### SBS18

The accumulation of SBS18 burden is asscoicated with age, but shows no difference across the three sections.
```{r SBS18, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS18  ~ age , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Coeliac history does not affect SBS18 burden
lmm.coeliac.burden <- lme(SBS18 ~ age + condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS18 burden
lmm.region.burden <- lme(SBS18 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS18,df_regression$age, method="pearson")

tmp=data.frame(condition=c(16,3))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = SBS18))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope =  fixed.m1['age',])+
  geom_ribbon(aes(ymin = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','lower'], ymax = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','upper']), alpha = 0.1)+
  labs(y='SBS18 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

### SBS2/13

The accumulation of SBS2 burden has a non-linear relationship with age.
```{r SBS2, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS2  ~ age, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")


# Coeliac history does not affect SBS2 burden
lmm.coeliac.burden <- lme(SBS2 ~ age + condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS2 burden
lmm.region.burden <- lme(SBS2 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS2,df_regression$age, method="pearson")

tmp=data.frame(condition=c(16,3))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y =  SBS2))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  labs(y='SBS2 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

The accumulation of SBS13 burden has a non-linear relationship with age.
```{r SBS13, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS13  ~ age, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Coeliac history does not affect SBS13 burden
lmm.coeliac.burden <- lme(SBS13 ~ age + condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS13 burden
lmm.region.burden <- lme(SBS13 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS13,df_regression$age, method="pearson")

tmp=data.frame(condition=c(16,3))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y =  SBS13))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  labs(y='SBS13 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

APOBEC burden is not correlated with reversetranposition events.
```{r}
lmm.APOBEC<- lme((SBS2+SBS13) ~ age + retrotransposition +region +condition +gender, random=list(patient = pdDiag(form = ~ age-1)), data = df_regression, method = 'ML') 
summary(lmm.APOBEC)
```

Estimate the coeliac effect on SBS:
```{r coeliac_SBS, fig.width = 6, fig.height=3, warning=FALSE}
lmm.sbs.total.coeliac.effect <- lme(sbs_count_adj ~ age:region + age:condition, random = list(patient = pdDiag(form = ~ age - 1)), data = df_regression, method = "ML")
lmm.sbs1.coeliac.effect <- lme(SBS1 ~ age:region + age:condition -1 , random = list(patient = pdDiag(form = ~ age - 1)), data = df_regression, method = "ML")
lmm.sbs5.coeliac.effect <- lme(SBS5 ~ age:region + age:condition, random = list(patient = pdDiag(form = ~ age - 1)), data = df_regression, method = "ML")
lmm.sbs18.coeliac.effect <- lme(SBS18 ~ age:region + age:condition, random = list(patient = pdDiag(form = ~ age - 1)), data = df_regression, method = "ML")
lmm.apobec.coeliac.effect <- lme((SBS2+SBS13) ~ age:region + age:condition, random = list(patient = pdDiag(form = ~ age - 1)), data = df_regression, method = "ML")

summary(lmm.sbs1.coeliac.effect)

df2 = data.frame(matrix(nrow=10,ncol=5))
colnames(df2) = c('rate','type','group','lower_95_CI','upper_95_CI')
df2[1,] = c(intervals(lmm.sbs.total.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','est.'],'Total SBS','Age effect',intervals(lmm.sbs.total.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'],
            intervals(lmm.sbs.total.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','upper'])
df2[2,] = c(intervals(lmm.sbs.total.coeliac.effect, which = "fixed")[["fixed"]]['age:conditionCoeliac','est.'],'Total SBS','Disease effect',intervals(lmm.sbs.total.coeliac.effect, which = "fixed")[["fixed"]]['age:conditionCoeliac','lower'],
            intervals(lmm.sbs.total.coeliac.effect, which = "fixed")[["fixed"]]['age:conditionCoeliac','upper'])
df2[3,] = c(intervals(lmm.sbs1.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','est.'],'SBS1','Age effect',intervals(lmm.sbs1.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'],
            intervals(lmm.sbs1.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','upper'])
df2[4,] = c(intervals(lmm.sbs1.coeliac.effect, which = "fixed")[["fixed"]]['age:conditionCoeliac','est.'],'SBS1','Disease effect',intervals(lmm.sbs1.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','lower'],
            intervals(lmm.sbs1.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','upper'])
df2[5,] = c(intervals(lmm.sbs5.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','est.'],'SBS5','Age effect',intervals(lmm.sbs5.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'],
            intervals(lmm.sbs5.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','upper'])
df2[6,] = c(intervals(lmm.sbs5.coeliac.effect, which = "fixed")[["fixed"]]['age:conditionCoeliac','est.'],'SBS5','Disease effect',intervals(lmm.sbs5.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','lower'],
            intervals(lmm.sbs5.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','upper'])
df2[7,] = c(intervals(lmm.sbs18.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','est.'],'SBS18','Age effect',intervals(lmm.sbs18.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'],
            intervals(lmm.sbs18.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','upper'])
df2[8,] = c(intervals(lmm.sbs18.coeliac.effect, which = "fixed")[["fixed"]]['age:conditionCoeliac','est.'],'SBS18','Disease effect',intervals(lmm.sbs18.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','lower'],
            intervals(lmm.sbs18.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','upper'])
df2[9,] = c(intervals(lmm.apobec.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','est.'],'SBS2/13','Age effect',intervals(lmm.apobec.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'],
            intervals(lmm.apobec.coeliac.effect, which = "fixed")[["fixed"]]['age:regionDuodenum','upper'])
df2[10,] = c(intervals(lmm.apobec.coeliac.effect, which = "fixed")[["fixed"]]['age:conditionCoeliac','est.'],'SBS2/13','Disease effect',intervals(lmm.apobec.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','lower'],
            intervals(lmm.apobec.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','upper'])


df2$rate = round(as.numeric(df2$rate), digit=2)
df2$lower_95_CI = round(as.numeric(df2$lower_95_CI), digit=2)
df2$upper_95_CI= round(as.numeric(df2$upper_95_CI), digit=2)
df2$type = factor(df2$type,levels = c('SBS2/13','SBS18','SBS5','SBS1','Total SBS'))
ggplot(df2[3:10,], aes(x=rate, y=type, group=group,color=group))+geom_point()+geom_errorbar(aes(xmin=lower_95_CI, xmax=upper_95_CI),width=.1)+
  theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  geom_vline(xintercept=0, linetype="dashed", color = "black")+
  labs(x='Substitutions per year',y=NULL)
```


Estimate the coeliac effect on Indels:
```{r coeliac_Indel, fig.width = 6, fig.height=3, warning=FALSE}
lmm.id.coeliac.effect <- lme( indel_count_adj  ~ age:region + age:condition, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression, method = "ML")
summary(lmm.id.coeliac.effect)
df3 = data.frame(matrix(nrow=2,ncol=5))
colnames(df3) = c('rate','type','group','lower_95_CI','upper_95_CI')
df3[1,] = c(intervals(lmm.id.coeliac.effect , which = "fixed")[["fixed"]]['age:regionDuodenum','est.'],'ID','Age effect',intervals(lmm.id.coeliac.effect , which = "fixed")[["fixed"]]['age:regionDuodenum','lower'],
            intervals(lmm.id.coeliac.effect , which = "fixed")[["fixed"]]['age:regionDuodenum','upper'])
df3[2,] = c(intervals(lmm.id.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','est.'],'ID','Disease effect',intervals(lmm.id.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','lower'],
            intervals(lmm.id.coeliac.effect , which = "fixed")[["fixed"]]['age:conditionCoeliac','upper'])
df3$rate = round(as.numeric(df3$rate), digit=2)
df3$lower_95_CI = round(as.numeric(df3$lower_95_CI), digit=2)
df3$upper_95_CI= round(as.numeric(df3$upper_95_CI), digit=2)
ggplot(df3, aes(x=rate, y=type, group=group,color=group))+geom_point()+geom_errorbar(aes(xmin=lower_95_CI, xmax=upper_95_CI),width=.03)+
  theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  geom_vline(xintercept=0, linetype="dashed", color = "black")+
  labs(x='Indels per year',y=NULL)

```
