---
title: "Mutational burden in normal human small intestine"
author: "Yichen Wang"
output: 
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=70), tidy=TRUE,warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = '/Users/yw2/OneDrive\ -\ University\ of\ Cambridge/PhD')
```

This analysis aims to assess 1) the rate of SBS and ID mutations, 2) the presence and the rate of mutaional burden of mutational sigantures in normal small intestine.

We build linear mixed models for estimating mutation rates. Model parameters encoded are: age, biopsy region (duodenum, jejunum, ileum) , celieac condition (having a coeliac history or not) and  patient. We have corrected for coverage by normalising ID and SBS mutational burden according to the sensitivity of detection in each samples before modelling, therfore we do not include it as a parameter for modelling.

## 1.Overview of dataset


```{r loading, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(nlme)
library(tidyverse)

# Load data
data=read.table('./data/stat_summary.txt',header=T,stringsAsFactors = F)
exposure_matrix_crypts = read.table("./data/sigs/exposure_matrix_crypts.txt",header=T,stringsAsFactors = F)
  

exposure_matrix_crypts = exposure_matrix_crypts[data$sample,]
data$SBS1 <- data$sbs_count_adj*exposure_matrix_crypts$SBS1
data$SBS5 <- data$sbs_count_adj*exposure_matrix_crypts$SBS5
data$SBS2 <- data$sbs_count_adj*exposure_matrix_crypts$SBS2
data$SBS13 <- data$sbs_count_adj*exposure_matrix_crypts$SBS13
data$SBS18 <- data$sbs_count_adj*exposure_matrix_crypts$SBS18
data$SBS88 <- data$sbs_count_adj*exposure_matrix_crypts$SBS88
data$SBS35 <- data$sbs_count_adj*exposure_matrix_crypts$SBS35
data$SBS17b <- data$sbs_count_adj*exposure_matrix_crypts$SBS17b
data$SBS40 <- data$sbs_count_adj*exposure_matrix_crypts$SBS40
data$SBS41 <- data$sbs_count_adj*exposure_matrix_crypts$SBS41

write.table(data,'./data/stat_summary.txt', quote=F, col.names = T, row.names = F)

# Exclude Brunner's glands
data<-data[which(data$ref=='Crypt'),]
table(data$patient)
dim(data)


df_regression = data
# Only includes samples with >15 coverage
df_regression<-df_regression[which(df_regression$coverage>15),]
# Exclude the individual with extensive chemotherapy and keep the one with little of it
df_regression<-df_regression[which(df_regression$patient != 'PD43853'),]
df_regression$condition[which(df_regression$patient == 'PD28690')] = "Normal"
# Exclude the two unusual cases
df_regression<-df_regression[! df_regression$sample %in% c('PD46565c_lo0009','PD43851j_P52_DDM_E2'),]
rownames(df_regression)=df_regression$sample
dim(df_regression)

```


```{r}
# Mean coverage
mean(data$coverage)

# Load SBS matrix
sbs<-read.table('./data/sbs/sbs_on_branch.txt',check.names = F)
# Exclude Brunner's glands
sbs=sbs[!(rownames(sbs) %in% c('PD43851_1','PD43851_4','PD43851_14')),]
# Total SBS mutations
sum(sbs)


# Load Indel matrix
indel<-read.table('./data/indel/indel_on_branch.txt',check.names = F)
# Exclude Brunner's glands
indel=indel[!(rownames(indel) %in% c('PD43851_1','PD43851_4','PD43851_14')),]
# Total IDmutations
sum(indel)
```

## 2. Hypothesis testing of linear mixed effects models
### Substitutions

First, we set up a linear mixed model with age as a fixed effect and patient as random effect.
```{r, warning=FALSE}
lmm.sbs.age <- lme(sbs_count_adj ~ age , random=list(patient = pdDiag(form = ~ age -1)), data = df_regression, method = 'ML') 
```

Coeliac history does not affect single-base substitution burdens.
```{r, warning=FALSE}
# Include coeliac history as a fixed effect does not improve the fitness of model.
lmm.sbs.coeliac.burden <- lme( sbs_count_adj  ~ age:condition + condition, 
                               random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.burden,lmm.sbs.age)
```

Coeliac history does not affect between-patient or within-patient heterogeneity of single-base substitution burdens.
```{r, warning=FALSE}
# Whether coeliac history affect inter-patient variation of single base substituions
lmm.sbs.coeliac.var.inter <- lme( sbs_count_adj  ~ age , random = list(patient = pdDiag(form = ~ age+condition-1)),  data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.var.inter, lmm.sbs.age)

# Whether coeliac history affect within-patient variation of single base substituions
lmm.sbs.coeliac.var.intra <- lme( sbs_count_adj  ~ age , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression, method = "ML")
anova(lmm.sbs.coeliac.var.intra, lmm.sbs.age)
```

Alyhough the p-value from ANNOVA test is significant, after careful investigation into the raw data, this variation in between-patient heterogeneity and heteroscedasticity for coeliac/normal patients was largely introduced by one single patient, PD46565. After removing this patient, distinguishing the two groups does not provide a better fit.

```{r, warning=FALSE}

lmm.sbs.age <- lme(sbs_count_adj ~ age , random=list(patient = pdDiag(form = ~ age -1)), data = df_regression[which(df_regression$patient != 'PD46565'),], method = 'ML') 
lmm.sbs.coeliac.var.inter <- lme(sbs_count_adj  ~ age , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ),  data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.sbs.coeliac.var.inter, lmm.sbs.age)

```

Biopsy region does not affect between-patient heterogeneity and heteroscedasticity. 
```{r warning=FALSE}
lmm.sbs.age <- lme(sbs_count_adj ~ age , random=list(patient = pdDiag(form = ~ age -1)), data = df_regression, method = 'ML') 
# Whether biopsy region affect inter-patient variation of single base substituions
lmm.sbs.region.var.inter<- lme( sbs_count_adj  ~ age, 
                                random = list(patient = pdDiag(form = ~ age+region -1)), data = df_regression, method = "ML")
anova(lmm.sbs.age,lmm.sbs.region.var.inter)

# Whether biopsy region affect within-patient variation of single base substituions
lmm.sbs.region.var.intra<- lme( sbs_count_adj  ~ age, 
                                random = list(patient = pdDiag(form = ~ age -1)), weights = varIdent(form = ~1 | region), data = df_regression, method = "ML")
anova(lmm.sbs.age,lmm.sbs.region.var.intra)

# The effect no longer exist after removing the outlier
lmm.sbs.age <- lme(sbs_count_adj ~ age , random=list(patient = pdDiag(form = ~ age -1)), data = df_regression[which(df_regression$patient != 'PD46565'),], method = 'ML') 
lmm.sbs.region.var.intra<- lme( sbs_count_adj  ~ age, 
                                random = list(patient = pdDiag(form = ~ age -1)), weights = varIdent(form = ~1 | region), data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.sbs.age,lmm.sbs.region.var.intra)



```

Allowing different mutaton rate for duodenum/jejunum/ileum improves the fitness of the model.
```{r warning=FALSE}

lmm.sbs.region.slope<- lme( sbs_count_adj  ~ age:region, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")
lmm.sbs.age <- lme(sbs_count_adj ~ age , 
                   random=list(patient = pdDiag(form = ~ age -1)), data = df_regression, method = 'ML') 
anova(lmm.sbs.region.slope,lmm.sbs.age)
```

Allowing different intercept for duodenum/jejunum/ileum significantly won't further improve the fitness of the model.
```{r warning=FALSE}

lmm.sbs.region.slope.intercept<- lme( sbs_count_adj  ~ age:region + region, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")
anova(lmm.sbs.region.slope,lmm.sbs.region.slope.intercept)
```

Therefore, our final model for SBS burden includes both age and biopsy region as fixed effects, and patient as a random effect.
```{r SBS_plot, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lmm.sbs.region.slope
summary(lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")


tmp=data.frame(condition=c(3,16))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = sbs_count_adj))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope = fixed.m1['age:regionDuodenum',],colour='#F8766D')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'], ymax = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','upper']),fill='#F8766D', alpha = 0.1)+
  geom_abline(intercept = (fixed.m1[1,]+fixed.m1['age:regionIleum',]), slope = fixed.m1['age:regionIleum',],colour='#00B81F')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionIleum',]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionIleum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','upper']), fill='#00B81F',alpha = 0.1)+
  geom_abline(intercept = fixed.m1[1,]+fixed.m1['age:regionJejunum',], slope = fixed.m1['age:regionJejunum',],colour='#00A5FF')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','upper']), fill='#00A5FF',alpha = 0.1)+
  labs(y='Substitutions / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

### Indels
As above, we set up a linear mixed model for Indels with age as a fixed effect and patient as random effect.
```{r warning=FALSE}
lmm.id.age <- lme(indel_count_adj ~ age, random=list(patient = pdDiag(form = ~ age-1)), data = df_regression, method = 'ML') 
```

Adding seperate mutaion rate for duodenum/jejunum/ileum significantly improves the fitness of the model.
```{r region indel, warning=FALSE}
# Whether biopsy regions affect Indel burden via slope
lmm.id.region.slope <- lme(indel_count_adj  ~ age:region, 
                           random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression, method = "ML")
anova(lmm.id.region.slope,lmm.id.age)
```

Allowing different intercepts for different sections won't further improve the model.
```{r}
# Whether biopsy regions affect Indel burden via slope and intercept
lmm.id.region.slope.intercept <- lme(indel_count_adj  ~ age:region + region, 
                                     random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression, method = "ML")
anova(lmm.id.region.slope.intercept,lmm.id.region.slope)
```

No significant inter-patient variation is caused by different biopsy sections, but adding different intra-patient variation can improve the fitness of the model.
```{r}
# Biopsy region does not affect between-patient variation of Indels
lmm.id.region.slope.var.inter <- lme( indel_count_adj  ~ age:region , 
                                      random = list(patient = pdDiag(form = ~ age+region-1)),  data = df_regression, method = "ML")
anova(lmm.id.region.slope.var.inter, lmm.id.region.slope)

# Biopsy region affects within-patient variation of Indels.The effect exist after removing possible outliers.
lmm.id.region.slope.var.intra <- lme( indel_count_adj  ~ age:region , 
                                      random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | region ), data = df_regression, method = "ML")
anova(lmm.id.region.slope.var.intra, lmm.id.region.slope)


lmm.id.region.slope <- lme(indel_count_adj  ~ age:region, 
                           random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
lmm.id.region.slope.var.intra <- lme( indel_count_adj  ~ age:region , 
                                      random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | region ), data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.id.region.slope.var.intra, lmm.id.region.slope)

```


Coeliac history does not affect Indel burdens, but affects intra-patient variation.
```{r coeliac indel, warning=FALSE}

# Coeliac history does not affect Indel burdens
lmm.id.region.slope <- lme(indel_count_adj  ~ age:region , 
                           random = list(patient = pdDiag(form = ~ age-1)),  data = df_regression, method = "ML")
lmm.id.region.burden.coeliac.burden <- lme(indel_count_adj  ~ age:region + age:condition +condition, 
                                           random = list(patient = pdDiag(form = ~ age-1)) ,data = df_regression, method = "ML")
anova(lmm.id.region.burden.coeliac.burden, lmm.id.region.slope)


# Coeliac history does not affect inter-patient variation of Indels
lmm.id.region.slope.coeliac.var.inter <- lme( indel_count_adj  ~ age:region , 
                                              random = list(patient = pdDiag(form = ~ age+condition-1)),  weights = varIdent(form = ~1 | region ), data = df_regression, method = "ML")
lmm.id.region.slope.var.intra <- lme( indel_count_adj  ~ age:region , 
                                      random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | region ), data = df_regression, method = "ML")
anova(lmm.id.region.slope.coeliac.var.inter, lmm.id.region.slope.var.intra)


# Coeliac history affects intra-patient variation of Indels. The effect exist after removing possible outliers.
lmm.id.region.slope.coeliac.var.intra <- lme( indel_count_adj  ~ age:region , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition*region ),  data = df_regression, method = "ML")
lmm.id.region.slope.var.intra <- lme( indel_count_adj  ~ age:region ,  random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | region ), data = df_regression, method = "ML")
anova(lmm.id.region.slope.coeliac.var.intra, lmm.id.region.slope.var.intra)

lmm.id.region.slope.var.intra.coeliac.var.intra<- lme( indel_count_adj  ~ age:region , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition*region ),  data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
lmm.id.region.slope.var.intra <- lme( indel_count_adj  ~ age:region , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | region ), data = df_regression[which(df_regression$patient != 'PD46565'),], method = "ML")
anova(lmm.id.region.slope.var.intra.coeliac.var.intra, lmm.id.region.slope.var.intra)
```

Taking coeliac history into consideration, we can have a reasonabely good model without adding biospy regions as a factor affecting intra-patient variation.
```{r}
lmm.id.region.slope.var.intra.coeliac.var.intra <- lme( indel_count_adj  ~ age:region ,  random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition*region ),  data = df_regression, method = "ML")
lmm.id.region.slope.coeliac.var.intra <- lme( indel_count_adj  ~ age:region , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 | condition ), data = df_regression, method = "ML")
anova(lmm.id.region.slope.var.intra.coeliac.var.intra, lmm.id.region.slope.coeliac.var.intra)
```

Therefore, our final model for ID burden includes both age and biopsy region as fixed effects, patient as a random effect, and use different covariance matrix for individual with/without a coeliac history.

```{r Indel_plot, fig.width = 6, fig.height=3, warning=FALSE}

lmm <-lmm.id.region.slope.coeliac.var.intra
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")


tmp=data.frame(condition=c(3,16))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = indel_count_adj))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope = fixed.m1['age:regionDuodenum',],colour='#F8766D')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'], ymax = fixed.m1[1,]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','upper']),fill='#F8766D', alpha = 0.1)+
  geom_abline(intercept = (fixed.m1[1,]+fixed.m1['age:regionIleum',]), slope = fixed.m1['age:regionIleum',],colour='#00B81F')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionIleum',]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionIleum',]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','upper']), fill='#00B81F',alpha = 0.1)+
  geom_abline(intercept = fixed.m1[1,]+fixed.m1['age:regionJejunum',], slope = fixed.m1['age:regionJejunum',],colour='#00A5FF')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+
                    age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','upper']), fill='#00A5FF',alpha = 0.1)+
  labs(y='Indels / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```


## 3. Mosiac plot for all signatures
```{r Mosiac_plot,fig.width = 10, fig.height=4}
df=exposure_matrix_crypts[rownames(exposure_matrix_crypts) %in% data$sample,]
df$patient = rownames(df)
df$age =data$age
df$patient = fct_reorder(df$patient, df$age, min)

test <- gather(df, E1, E2, -patient, -age)
test$E1<-factor(test$E1,levels = c('SBS1','SBS5','SBS18','SBS2','SBS13','SBS88','SBS35','SBS40','SBS41','SBS17b'))
test$group = 'Normal'
test$group[test$patient %in% data$sample[data$condition == 'Coeliac']] = 'Coeliac'

sig_order=1:10
names(sig_order)=c("SBS1","SBS5","SBS18","SBS2","SBS13","SBS88","SBS35","SBS40",'SBS41',"SBS17b")
final_sigs=t(read.table("./data/sigs/final_sigs.txt",check.names=FALSE))

getPalette = colorRampPalette(brewer.pal(8, "Set3"))
all_cols=getPalette(8)
all_cols=c(all_cols,"firebrick","magenta")
final_sigs = final_sigs[names(sort(sig_order[rownames(final_sigs)])),]
names(all_cols)=rownames(final_sigs)

ggplot(test, aes(x = patient, y = E2, fill = E1)) + geom_bar(stat = "identity") +scale_fill_manual(values=all_cols)+
  theme_bw()+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(), panel.grid=element_blank(),panel.border=element_blank())+
  labs(x = "Samples", y = "Proportion", fill = "Sigantures")+facet_grid(cols = vars(group),scales = "free_x", space = "free")

```

## 4. The presence of APOBEC signatures SBS2/13 in the small intestine
```{r small_bowel_vs_colon, fig.width = 2.5, fig.height = 3}
exposure_matrix_all = read.table("./data/sigs/exposure_matrix_with_colon.txt",header=T,stringsAsFactors = F)
input_for_hdp<-read.table('./data/input_for_hdp.txt',check.names = F)
input_for_hdp=input_for_hdp[apply(input_for_hdp,1,sum)>50,]
input_for_hdp=input_for_hdp[!(rownames(input_for_hdp) %in% c('PD43851_1','PD43851_4','PD43851_14')),]

apobec_table = data.frame(matrix(ncol = 2, nrow = length(unique(data$patient))))
rownames(apobec_table)=unique(data$patient)
colnames(apobec_table)=c('Positive','Negative')
for (patient in unique(data$patient)){
  apobec_table[patient,'Positive']=sum(sum(data[data$patient==patient,'SBS2']+data[data$patient==patient,'SBS13']>0.05*data[data$patient==patient,'sbs_count_adj']))
  apobec_table[patient,'Negative']=sum(data$patient==patient)-apobec_table[patient,'Positive']
}

APOBEC_matrix_all=exposure_matrix_all[,c('SBS2','SBS13')]
APOBEC_matrix_all$sum = APOBEC_matrix_all$SBS2 + APOBEC_matrix_all$SBS13
APOBEC_matrix_all$sum[APOBEC_matrix_all$sum<0.05] = 0
APOBEC_matrix_all$group = 'Colon'
APOBEC_matrix_all$group[rownames(exposure_matrix_all) %in% rownames(input_for_hdp)] = 'Small intestine'
APOBEC_matrix_all$branch = rownames(APOBEC_matrix_all)

df=data.frame(matrix(ncol = 3, nrow = 2))
colnames(df) <- c('region', 'pos', 'neg')
df$region = c('Small intestine','Colon')
df$pos = c(sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Small intestine'] > 0),sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Colon'] > 0))
df$neg = c(sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Small intestine'] == 0),sum(APOBEC_matrix_all$sum[APOBEC_matrix_all$group == 'Colon'] == 0))


df[1,2:3] = df[1,2:3]/sum(df[1,2:3])
df[2,2:3] = df[2,2:3]/sum(df[2,2:3])

test <- gather(df, E1, E2, -region) 
test$E1 <- factor(test$E1,levels = c('pos','neg'))


ggplot(test[test$E1=='pos',]) +
  geom_bar(aes(x = region, y = E2, fill = E1),
           stat = "identity") +scale_fill_brewer(palette="Paired",labels=c('APOBEC positive','APOBEC negative'))+theme_bw()+theme(panel.grid=element_blank(),legend.position="none",legend.title=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  labs(x = "Location", y = "APOBEC frequency")

```

## 5. Mutational burden for common mutational signatures in the normal small intestine
We looked into the accumulation of mutational buren across the corhort for SBS1, SBS5, SBS18 and SBS2/13. Other sporadic signatures are too rare for conducting meaningful statistical analysis.

### SBS1

The accumulation of SBS1 burden is asscoicated with age and differ in diifferent biopsy regions.
```{r SBS1, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS1  ~ age -1, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Biopsy region affects SBS1 burden
lmm.region.burden <- lme(SBS1 ~ age:region -1, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)

# Coeliac history does not affect SBS1 burden
lmm.coeliac.burden <- lme(SBS1 ~ age:region+age:condition -1, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm.region.burden)

lmm <- lmm.region.burden
fixed.m1 <- data.frame(fixef(lmm))

intervals(lmm, which = "fixed")
cor(df_regression$SBS1,df_regression$age, method="pearson")


tmp=data.frame(condition=c(3,16))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = SBS1))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope = fixed.m1['age:regionDuodenum',],colour='#F8766D')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','lower'], ymax = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionDuodenum','upper']),fill='#F8766D', alpha = 0.1)+
  geom_abline(intercept = (fixed.m1[1,]+fixed.m1['age:regionIleum',]), slope = fixed.m1['age:regionIleum',],colour='#00B81F')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionIleum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionIleum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionIleum','upper']), fill='#00B81F',alpha = 0.1)+
  geom_abline(intercept = fixed.m1[1,]+fixed.m1['age:regionJejunum',], slope = fixed.m1['age:regionJejunum',],colour='#00A5FF')+
  geom_ribbon(aes(ymin = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','lower'], ymax = fixed.m1[1,]+fixed.m1['age:regionJejunum',]+age*intervals(lmm, which = "fixed")[["fixed"]]['age:regionJejunum','upper']), fill='#00A5FF',alpha = 0.1)+
  labs(y='SBS1 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

### SBS5

The accumulation of SBS5 burden is asscoicated with age, but shows no difference across the three sections.
```{r SBS5, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS5  ~ age , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Coeliac history does not affect SBS5 burden
lmm.coeliac.burden <- lme(SBS5 ~ age:condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS5 burden
lmm.region.burden <- lme(SBS5 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS5,df_regression$age, method="pearson")

tmp=data.frame(condition=c(3,16))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = SBS5))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope = fixed.m1['age',])+
  geom_ribbon(aes(ymin = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','lower'], ymax = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','upper']), alpha = 0.1)+
  labs(y='SBS5 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))

```

### SBS18

The accumulation of SBS18 burden is asscoicated with age, but shows no difference across the three sections.
```{r SBS18, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS18  ~ age , random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Coeliac history does not affect SBS18 burden
lmm.coeliac.burden <- lme(SBS18 ~ age:condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS18 burden
lmm.region.burden <- lme(SBS18 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS18,df_regression$age, method="pearson")

tmp=data.frame(condition=c(3,16))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y = SBS18))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  geom_abline(intercept = fixed.m1[1,], slope = fixed.m1['age',])+
  geom_ribbon(aes(ymin = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','lower'], ymax = fixed.m1[1,]+age*intervals(lmm, which = "fixed")[["fixed"]]['age','upper']), alpha = 0.1)+
  labs(y='SBS18 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

### SBS2

The accumulation of SBS2 burden has a non-linear relationship with age.
```{r SBS2, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS2  ~ age, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Coeliac history does not affect SBS2 burden
lmm.coeliac.burden <- lme(SBS2 ~ age:condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS2 burden
lmm.region.burden <- lme(SBS2 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS2,df_regression$age, method="pearson")

tmp=data.frame(condition=c(3,16))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y =  SBS2))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  labs(y='SBS2 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```


### SBS13

The accumulation of SBS13 burden has a non-linear relationship with age.
```{r SBS13, warning=FALSE, fig.width = 6, fig.height=3}
lmm <- lme( SBS13  ~ age, random = list(patient = pdDiag(form = ~ age-1)), weights = varIdent(form = ~1 ), data = df_regression, method = "ML")

# Coeliac history does not affect SBS13 burden
lmm.coeliac.burden <- lme(SBS13 ~ age:condition, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.coeliac.burden, lmm)

# Biopsy region does not affect SBS13 burden
lmm.region.burden <- lme(SBS13 ~ age:region, random = list(patient = pdDiag(form = ~age - 1)), data = df_regression, method = "ML")
anova(lmm.region.burden, lmm)
fixed.m1 <- data.frame(fixef(lmm))
intervals(lmm, which = "fixed")
cor(df_regression$SBS13,df_regression$age, method="pearson")

tmp=data.frame(condition=c(3,16))
tmp_name = c('Coeliac','Normal')
rownames(tmp)=tmp_name

ggplot(data = df_regression, mapping = aes(x = age, y =  SBS13))+geom_point(aes(colour = region,fill=region,shape=condition)) +theme_bw()+theme(panel.grid=element_blank(),panel.border=element_blank(),axis.line=element_line(size=1,colour="black"))+
  scale_shape_manual(values = tmp$condition)+
  labs(y='SBS13 burden / crypt',x="Age (yrs)")+
  theme(title=element_text(size=18),
        axis.text.y=element_text(size=18,color="black"),
        axis.text.x=element_text(size=18,color="black"),
    legend.text=element_text(size=16))
```

